---
title: "DESeq"
format: html
---

# DESeq analysis

## Load environment and data

```{r}
library(tidyverse)
library(DESeq2)
library(fgsea)
library(msigdbr)
```

Check that colnames in count data and row names of metadata are in the same order

```{r}
all(metadata_df_77_sub$sample == colnames(gene_count_77_sub))
```

# DeSeq Condition

```{r}
dds_77_sub <- DESeqDataSetFromMatrix(
  countData = gene_count_77_sub, 
  colData = metadata_df_77_sub,
  design = ~ groups)
```

### PCA plot

converting counts to integer mode

```{r}
dds_77_sub
```

```{r}
pca_77_sub <- plotPCA(rlog(dds_77_sub), intgroup=c("groups"))
pca_77_sub
```

```{r}
ggsave("../fig/pca_77_sub.png", plot = pca_77_sub)
```

COMMENT ON THIS

### Testing for differential expression

```{r}
dds_77_sub <- DESeq(dds_77_sub)
```

```{r}
res_77_sub <- results(dds_77_sub)
```

How many genes were significant using an alpha of 0.05?

```{r}
n_sig_77_sub <- sum(res_77_sub$padj < 0.05, na.rm = TRUE)
n_sig_77_sub
```

```{r}
nrow(gene_count_77_sub)
```

Out of 67,186 genes, 40 showed significant differential expression between the two groups/conditions. We expect a 5% false discovery rate among the genes we call significant.

### MA-plot

```{r}
plotMA(res_77, ylim=c(-2,2))
dev.off()

MA_77_sub <- plotMA(res_77, ylim=c(-2,2))

```

```{r}
ggsave("../fig/MA_77_sub.png", plot = MA_77_sub)
png(filename = "../fig/MA_77_sub.png")

```

# DeSeq Condition&Gender

Confounders are external variables that are linked to both the independent and dependent variables in a study, distorting the true relationship between them and leading to a misleading conclusion. Let’s check if there is a gender confounder in the data

```{r}
dds_gender_77_sub <- DESeqDataSetFromMatrix(
  countData = gene_count_77_sub,
  colData = metadata_df_77_sub,
  design = ~ gender + groups)
```

```{r}
dds_gender_77_sub
```

```{r}
pca_77_gender_sub <- plotPCA(rlog(dds_gender_77_sub), intgroup=c("groups", "gender"))
pca_77_gender_sub
```

Save plot

```{r}
ggsave("../fig/pca_77_gender_sub.png", plot = pca_77_gender_sub)
```

COMMENT ON THIS

```{r}
dds_gender_77_sub <- DESeq(dds_gender_77_sub)
```

A co-factor is *any variable that explains expression variation*, even if it is balanced between conditions and therefore not a confounder.

```{r}
res_gender_77_sub <- results(dds_gender_77_sub,
                         contrast =c(
                          "groups",
                          "healthy",
                          "parkinson's")
                         )
```

How many genes were significant using an alpha of 0.05?\

```{r}
n_sig_gender_77_sub <- sum(res_gender_77_sub$padj < 0.05, na.rm = TRUE)
n_sig_gender_77_sub
```

What is the overlap between the differential analysis with and without modeling gender?\

```{r}
signif_77_sub <- rownames(res_77_sub)[!is.na(res_77_sub$padj) & res_77_sub$padj < 0.05]
signif_gender_77_sub <- rownames(res_gender_77_sub)[!is.na(res_gender_77_sub$padj) & res_gender_77_sub$padj < 0.05]

# Number of genes in common
n_common_77_sub <- length(intersect(signif_77_sub, signif_gender_77_sub))
n_common_77_sub
```

Gender looks like a modest but real co-factor.

The inclusion of gender in the model led to the identification of five more differentially expressed (DE) genes (45 total vs. 40), suggesting that gender is a factor that influences the expression of these few genes and should be retained in the model to reduce noise.

Eventhough the PCA does not totally agree by separating, the analysis showed that adjusting for gender helps attribute the condition effect more clearly.

```{r}
saveRDS(res_gender_77_sub, file = "../data/deseq2_results_77.rds")
```

# DeSeq Condition&Gender&Age

```{r}
dds_gender_age_77_sub <- DESeqDataSetFromMatrix(
  countData = gene_count_77_sub,
  colData = metadata_df_77_sub,
  design = ~  groups + age + gender)
```

```{r}
dds_gender_age_77_sub <- DESeq(dds_gender_age_77_sub)

```

```{r}
pca_gender_age_77_sub <- plotPCA(rlog(dds_gender_age_77_sub), intgroup=c("groups","age"))
pca_gender_age_77_sub
```

Save the plot

```{r}
ggsave("../fig/pca_gender_age_77_sub.png", plot = pca_gender_age_77_sub)
```

```{r}
res_gender_age_77_sub <- results(dds_gender_age_77_sub,
                         contrast =c(
                          "groups",
                          "healthy",
                          "parkinson's")
                         )

# How many genes were significant using an alpha of 0.05?
n_sig_gender_age_77_sub <- sum(res_gender_age_77_sub$padj < 0.05, na.rm = TRUE)
cat("number of significant genes when modelling with gender and age:", n_sig_gender_age_77_sub, "\n")

# What is the overlap between the differential analysis with and without modeling age?
signif_gender_77_sub <- rownames(res_77_sub)[!is.na(res_gender_77_sub$padj) & res_gender_77_sub$padj < 0.05]
signif_gender_age_77_sub <- rownames(res_gender_age_77_sub)[!is.na(res_gender_age_77_sub$padj) & res_gender_age_77_sub$padj < 0.05]

# Number of genes in common
n_common_77_sub <- length(intersect(signif_gender_77_sub, signif_gender_age_77_sub))
cat("number of overlapping significant genes when modelling with gender and age, and just gender: ",n_common_77_sub)
```

Save the results

```{r}
saveRDS(res_gender_age_77_sub, file = "../data/deseq2_results_77_age_sub.rds")
```

Conclusion: age is not a confounder!

### MA-plot inlc gender + age

```{r}
plotMA(res_gender_age_77, ylim=c(-2,2))
dev.off()

MA_gender_age_77 <- plotMA(res_gender_age_77, ylim=c(-2,2))
```

```{r}
ggsave("../fig/MA_gender_age_77.png", plot = MA_gender_age_77)
png(filename = "../fig/MA_gender_age_77.png")

```

# Gene-set enrichment analysis

!!Not with right results yet!!

TODO: Do with res_gender_age_77

## FSC

```{r}
BP_df = msigdbr(species = "human", category = "C5", subcategory = "BP")
BP_list = split(BP_df$ensembl_gene, BP_df$gs_name)
```

```{r}
# Create the ranked list of gene-level statistics
myStats_77 <- res_gender_77[,"stat"]
names(myStats_77) <- rownames(res_gender_77)

# Remove the ensembl version in the names
clean_names_77 <- sub("\\.\\d+$", "", names(myStats_77))
names(myStats_77) <- clean_names_77

myStats_77 <- myStats_77[!is.na(myStats_77)]
myStats_77 <- sort(myStats_77, decreasing = TRUE)
```

```{r}
fg_77 <- fgseaMultilevel(pathways = BP_list, stats = myStats_77, minSize = 15, maxSize = 500, scoreType = "std")
```

```{r}
fg_sig_77 <- fg_77 |> arrange(padj) |> filter(padj < 0.05)

nrow(fg_sig_77)
```

```{r}
fg_sig_77[1:10]
```

COMMENT ON THIS

```{r}
chosen_pathway_77 <- fg_sig_77$pathway[1]
enrichment_77_direct <- plotEnrichment(BP_list[[chosen_pathway_77]], myStats_77) + ggplot2::labs(title = paste0("Enrichment: ", chosen_pathway_77))
enrichment_77_direct
```

```{r}
ggsave("../fig/enrichment_77_direct.png", plot = enrichment_77_direct)
```

```{r}
leading_77 <- fg_sig_77$leadingEdge[[1]]
first_genes_77 <- head(leading_77, 3)
first_genes_77
```

## FSC including age

```{r}
# Create the ranked list of gene-level statistics
myStats_77 <- res_gender_age_77[,"stat"]
names(myStats_77) <- rownames(res_gender_age_77)

# Remove the ensembl version in the names
clean_names_77 <- sub("\\.\\d+$", "", names(myStats_77))
names(myStats_77) <- clean_names_77

myStats_77 <- myStats_77[!is.na(myStats_77)]
myStats_77 <- sort(myStats_77, decreasing = TRUE)
```

```{r}
fg_77 <- fgseaMultilevel(pathways = BP_list, stats = myStats_77, minSize = 15, maxSize = 500, scoreType = "std")
```

```{r}
fg_sig_77 <- fg_77 |> arrange(padj) |> filter(padj < 0.05)

nrow(fg_sig_77)
```

```{r}
fg_sig_77[1:100, c("pathway", "padj", "NES")]
```

```{r}
enrichment_77_direct <- plotEnrichment(BP_list[[chosen_pathway_77]], myStats_77)
enrichment_77_direct <- enrichment_77_direct + ggplot2::labs(title = paste0("Enrichment: ", chosen_pathway_77))
enrichment_77_direct

# save plot
ggsave("../fig/enrichment_77_direct_age.png", plot = enrichment_77_direct)
```

COMMENT ON THIS

## FSC as non-directional including age

```{r}
chosen_pathway_77 <- fg_sig_77$pathway[1]
enrichment_77_direct <- plotEnrichment(BP_list[[chosen_pathway_77]], myStats_77) +ggplot2::labs(title = paste0("Enrichment: ", chosen_pathway_77))
enrichment_77_direct
```

```{r}
stats_abs_77 <- abs(myStats_77)
fg_nondir_77 <- fgseaMultilevel(pathways = BP_list, stats = stats_abs_77,
minSize = 15, maxSize = 500, scoreType = "pos")
```

```{r}
fg_nondir_77 <- fg_nondir_77[order(fg_nondir_77$padj), ]
fg_nondir_77[1:10]
```

COMMENT ON THIS

```{r}
enrichment_77_non_direct <- plotEnrichment(BP_list[[chosen_pathway_77]], stats_abs_77)
enrichment_77_non_direct <- enrichment_77_non_direct + ggplot2::labs(title = paste0("Enrichment: ", chosen_pathway_77))

enrichment_77_non_direct
# save plot
ggsave("../fig/enrichment_77_non_direct_age.png", plot = enrichment_77_non_direct)
```
