---
title: "DESeq"
format: html
---

# DESeq analysis

## Load environment and data

```{r}
library(tidyverse)
library(DESeq2)
library(fgsea)
library(msigdbr)
```

Check that colnames in count data and row names of metadata are in the same order

```{r}
all(metadata_df_77$sample == colnames(gene_count_77))
```

# DeSeq Condition

```{r}
dds_77 <- DESeqDataSetFromMatrix(
  countData = gene_count_77, 
  colData = metadata_df_77,
  design = ~ groups)
```

#### PCA plot

converting counts to integer mode

```{r}
dds_77
```

```{r}
pca_77 <- plotPCA(rlog(dds_77), intgroup=c("groups"))
pca_77
```

```{r}
ggsave("../fig/pca_77.png", plot = pca_77)
```

COMMENT ON THIS

#### Testing for differential expression

```{r}
dds_77 <- DESeq(dds_77)
```

```{r}
res_77 <- results(dds_77)
```

How many genes were significant using an alpha of 0.05?

```{r}
n_sig_77 <- sum(res_77$padj < 0.05, na.rm = TRUE)
n_sig_77
```

```{r}
nrow(gene_count_77)
```

COMMENT ON THIS

# DeSeq Condition&Gender

Confounders are external variables that are linked to both the independent and dependent variables in a study, distorting the true relationship between them and leading to a misleading conclusion. Let’s check if there is a gender confounder in the data

```{r}
dds_gender_77 <- DESeqDataSetFromMatrix(
  countData = gene_count_77,
  colData = metadata_df_77,
  design = ~ gender + groups)
```

```{r}
dds_gender_77
```

```{r}
pca_77_gender <- plotPCA(rlog(dds_gender_77), intgroup=c("groups", "gender"))
pca_77_gender
```

Save plot

```{r}
ggsave("../fig/pca_77_gender.png", plot = pca_77_gender)
```

COMMENT ON THIS

```{r}
dds_gender_77 <- DESeq(dds_gender_77)
```

A co-factor is *any variable that explains expression variation*, even if it is balanced between conditions and therefore not a confounder.

```{r}
res_gender_77 <- results(dds_gender_77,
                         contrast =c(
                          "groups",
                          "healthy",
                          "parkinson's")
                         )
```

How many genes were significant using an alpha of 0.05?\

```{r}
n_sig_gender_77 <- sum(res_gender_77$padj < 0.05, na.rm = TRUE)
n_sig_gender_77
```

What is the overlap between the differential analysis with and without modeling gender?\

```{r}
signif_77 <- rownames(res_77)[!is.na(res_77$padj) & res_77$padj < 0.05]
signif_gender_77 <- rownames(res_gender_77)[!is.na(res_gender_77$padj) & res_gender_77$padj < 0.05]

# Number of genes in common
n_common_77 <- length(intersect(signif_77, signif_gender_77))
n_common_77
```

COMMENT ON THIS

```{r}
saveRDS(res_gender_77, file = "../data/deseq2_results_77.rds")
```

# DeSeq Condition&Gender&Age

```{r}
dds_gender_age_77 <- DESeqDataSetFromMatrix(
  countData = gene_count_77,
  colData = metadata_df_77,
  design = ~  groups + age + gender)
```

```{r}
dds_gender_age_77 <- DESeq(dds_gender_age_77)

```

```{r}
pca_gender_age_77 <- plotPCA(rlog(dds_gender_age_77), intgroup=c("groups","age"))
pca_gender_age_77
```

Save the plot

```{r}
ggsave("../fig/pca_gender_age_77.png", plot = pca_gender_age_77)
```

```{r}
res_gender_age_77 <- results(dds_gender_age_77,
                         contrast =c(
                          "groups",
                          "healthy",
                          "parkinson's")
                         )

# How many genes were significant using an alpha of 0.05?
n_sig_gender_age_77 <- sum(res_gender_age_77$padj < 0.05, na.rm = TRUE)
cat("number of significant genes when modelling with gender:", n_sig_gender_age_77, "\n")

# What is the overlap between the differential analysis with and without modeling age?
signif_gender_77 <- rownames(res_77)[!is.na(res_gender_77$padj) & res_gender_77$padj < 0.05]
signif_gender_age_77 <- rownames(res_gender_age_77)[!is.na(res_gender_age_77$padj) & res_gender_age_77$padj < 0.05]

# Number of genes in common
n_common_77 <- length(intersect(signif_gender_77, signif_gender_age_77))
cat("number of overlapping significant genes when modelling with gender and age, and just gender: ",n_common_77)
```

Save the results

```{r}
saveRDS(res_gender_age_77, file = "../data/deseq2_results_77_age.rds")
```

Conclusion: age is not a confounder!

# Gene-set enrichment analysis

!!Not with right results yet!!!

TODO: Do with res_gender_age_77

## FSC

```{r}
BP_df = msigdbr(species = "human", category = "C5", subcategory = "BP")
BP_list = split(BP_df$ensembl_gene, BP_df$gs_name)
```

```{r}
# Create the ranked list of gene-level statistics
myStats_77 <- res_gender_77[,"stat"]
names(myStats_77) <- rownames(res_gender_77)

# Remove the ensembl version in the names
clean_names_77 <- sub("\\.\\d+$", "", names(myStats_77))
names(myStats_77) <- clean_names_77

myStats_77 <- myStats_77[!is.na(myStats_77)]
myStats_77 <- sort(myStats_77, decreasing = TRUE)
```

```{r}
fg_77 <- fgseaMultilevel(pathways = BP_list, stats = myStats_77, minSize = 15, maxSize = 500, scoreType = "std")
```

```{r}
fg_sig_77 <- fg_77 |> arrange(padj) |> filter(padj < 0.05)

nrow(fg_sig_77)
```

```{r}
fg_sig_77[1:10]
```

COMMENT ON THIS

```{r}
chosen_pathway_77 <- fg_sig_77$pathway[1]
plotEnrichment(BP_list[[chosen_pathway_77]], myStats_77) +
enrichment_77_direct <- ggplot2::labs(title = paste0("Enrichment: ", chosen_pathway_77))
enrichment_77_direct
```

```{r}
ggsave("../fig/enrichment_77_direct.png", plot = enrichment_77_direct)
```

```{r}
leading_77 <- fg_sig_77$leadingEdge[[1]]
first_genes_77 <- head(leading_77, 3)
first_genes_77
```

COMMENT ON THIS

## FSC as non-directional

```{r}
stats_abs_77 <- abs(myStats_77)
fg_nondir_77 <- fgseaMultilevel(pathways = BP_list, stats = stats_abs_77,
minSize = 15, maxSize = 500, scoreType = "pos")
```

```{r}
fg_nondir_77 <- fg_nondir_77[order(fg_nondir_77$padj), ]
fg_nondir_77[1:10]
```

COMMENT ON THIS
