---
title: "DESeq"
format: html
---

### DESeq analysis

#### Load environment and data

```{r}
library(tidyverse)
library(DESeq2)
library(fgsea)
library(msigdbr)
```

##### Check that colnames in count data and row names of metadata are in the same order

```{r}
all(metadata_df_08$sample == colnames(gene_count_08))
```

```{r}
dds_08 <- DESeqDataSetFromMatrix(
  countData = gene_count_08, 
  colData = metadata_df_08,
  design = ~ groups)
```

#### PCA plot

converting counts to integer mode

```{r}
dds_08
```

```{r}
plotPCA(rlog(dds_08), intgroup=c("groups"))
```

#### Testing for differential expression

```{r}
dds_08 <- DESeq(dds_08)
```

```{r}
res_08 <- results(dds_08)
```

How many genes were significant using an alpha of 0.05?

```{r}
n_sig_08 <- sum(res_08$padj < 0.05, na.rm = TRUE)
n_sig_08
```

```{r}
nrow(gene_count_08)
```

Out of 67,186 genes showed significant differential expression between the two groups/conditions. We expect a 5% false discovery rate among the genes we call significant.

##### Correcting for confounders

Confounders are external variables that are linked to both the independent and dependent variables in a study, distorting the true relationship between them and leading to a misleading conclusion. Let’s check if there is a gender confounder in the data
